---
title: "Introduction to Psychometrics and Acale Sevelopment"
author: "Micah Gerhardt"
date: "11/12/2021"
output: powerpoint_presentation
---


## Introduction
-Psychometrics is a large field that deals with questionnaire development and measurement.
-This presentation is only a small part of what is involved and is focused more on the programming/statistical components
-This is also going to focus on pretty basic concepts

## What is Psychometrics?
-As a field psychometrics is concerned with measurement, testing, and scale development
-This includes concepts such as validity, reliability, and the theory of testing
-Two of the major methods used for scale development are: Factor Analysis and Item Response Theory 
-We will only cover factor analysis today

## History of Factor Analysis
-Developed in 1904 by Charles Spearman to support his hypothesis of intelligence

-Resulted in Spearman's g, a general factor of intelligence

-Intelligence would go on to become a central concept in Psychology 

-Factor analysis would be widely used in a variety of applications and is still one of the most used methods today

-Louis Thurstone would later develop multifactor models and discussed issues such as rotation and simple structure

## What is Factor Analysis?
-Factor analysis is a method of grouping measured variables together by shared variance
-Not enough time to get into the math but there are a number of resources (It is all matrix algebra)
-In modern day being able to do the matrix algebra by hand isn't that big of a deal. Commonly a 1 time assignment in factor analysis classes to do a simple factor analysis
-Factor analysis involves estimating latent variable(s) from manifest/measured variables 

## Latent Variables
-One can view there being two type of variables: latent variables and manifest or measured variables
-Manifest variables are what you are used to and can be directly measured (e.g., Height, blood pressure)
-Latent variables are thought only to exist at a latent level, in psychometrics this often means thoughts, patterns of thinking, mental state, etc.
-Conceptually latent variables exist within the individual and **cause** them to respond to questions in certain manners

## Latent Variables (continued)
-Every latent variable is thought to be normally distributed in the population
-You are likely already familiar with a number of latent variables
-As a side note when estimating these we typically set a mean of 0 and variance of 1

## What isn't Factor Analysis?
-**It is not a principal component analysis**
-The clearest difference is that in PCA the component is a linear combination of the variables
-As stated before factor analysis views the measured variables as resulting from the underlying latent factor
-People often gets PCA confused with factor analysis


## Factor Analysis and Psychometrics
-In psychometrics factor analysis is grouped into two types: Exploratory Factor Analysis (EFA) and Confirmatory Factor Analysis (CFA)
-You will also hear them called unrestricted factor analysis and restricted factor analysis
-The general process for scale development is:
-Write questions
-Get respondents
-Examine descriptives
-Conduct EFA
-Conduct CFA
-Adjust questions if needed and start again at get respondents step
-New scales typically need to go though this process several times

## Exploratory Factor Analysis
```{r, Descriptives}
#install.packages('psych', dependencies = T)
library(psych)
describe(Thurstone)
```

## Parallel Analysis
-Best method to determine number of factors
```{r, Parallel Analysis}
fa.parallel(Thurstone)
```
-Looks like 3 factor solution is best

## EFA
```{r, EFA}
fa(Thurstone, 3, rotate="varimax")
fa(Thurstone, 1, rotate="oblimin")
fa(Thurstone, 3, rotate="oblimin")
```

## Summary of EFA
-Normally I would copy and paste the MR1, MR2, and MR3 from the last EFA into excel and highlight the large values but in the interest of time I won't do that today
-In the interest of time we will go with the 3 factor oblimin solution
-Normally you name the factors
-In this case this is an old data set with three latent variables: Verbal Comprehension, Word Fluency, and Reasoning


## CFA
-Now that we have done the EFA we move onto the CFA
-To get out of rotational indeterminacy we restrict paths so that there is a unique best solution
-Will always fit worse than EFA

## Software
-Normally CFA are done in more specialized software
-psych package is installed on the server here
-The package, lavaan, I will use from this point on is not
-lavaan is a super useful package and comparable software is often $500+ a year for a personal license

## CFA code
-Take the EFA factor structure and make it a CFA with only the relevant paths
```{r, measurement model}
library(lavaan)

model <- "
Verbal Comprehension =~ Sentences + Vocabulary + Sent.Completion
Reasoning =~ First.Letters + Four.Letter.Words + Suffixes
Word Fluency =~ Letter.Series + Pedigrees + Letter.Group
"
fit <- cfa(model, sample.cov = Thurstone, sample.nobs = 213, estimator = "ML")
summary(fit)
```

## CFA model fit
-Have to evaluate model fit
```{r, model fit}
fitMeasures(fit)
```
-Generally want chi-square, a comparative fit index (CFI or TLI), and an absolute fit index (RMSEA)
-CFI or TLI want greater than .9 and prefer .95; RMSEA want below .06 if possible but below ~.09 is perhaps publishable (with great difficulty)
-Chi-square is overly sensitive and often would reject the model, not really used for model fit
-I would evaluate this model as being good

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

